---
title: "ezlimmaplot vignette"
author: "Jonathan Dreyfuss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ezlimmaplot vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
Welcome to **ezlimmaplot**. This package is for plotting bioinformatics results, especially those from the **ezlimma** package. Here, I'll load both packages, create some output with **ezlimma**, and then plot it with **ezlimmaplot**.

## Install package
See README.md.

## Load package
```{r lib}
library(ezlimma)
library(ezlimmaplot)
```

## Simulate data
To demonstrate the package on a dataset, I'll simulate log2 expression data with 100 genes and 6 samples in two groups, with the first gene up-regulated in the first group. This is the example data from **limma** `contrasts.fit`.
```{r sim}
set.seed(42)
M <- matrix(rnorm(100*6, sd=0.3), nrow=100, ncol=6)
dimnames(M) <- list(paste0("gene", 1:nrow(M)), paste0("sample", 1:ncol(M)))
grp <- rep(c("First3", "Last3"), each=3)
tissue <- rep(c("muscle", "liver"), times=3)
pheno <- data.frame(row.names = colnames(M), sample=colnames(M), grp=grp, tissue=tissue, stringsAsFactors = FALSE)
M[1,1:3] <- M[1,1:3] + 2
```

## Principal component analysis (PCA)
We can plot the samples by their first two principal components using the **ggplot2** package.
```{r pca}
ezpca(M, pheno, color="grp", name=NA)
```

We set `name=NA` to suppress plotting to a PNG file. We pass `color="grp"` via ... to **ggplot2** `aes_string`, where `"grp"` can be any of `colnames(pheno)}`. We could similarly have used `color="tissue"`. We can also facet by a phenotype column using the `facet` argument, which takes a formula.

```{r pca_facet}
ezpca(M, pheno, color="grp", facet = ". ~ tissue", name=NA)
```

## Gene level differential analysis
Assume we want to test if genes are changed in First3, in Last3, or in Last3 - First3. We can do this in two lines of code,
```{r lc}
contr.v <- c(First3="First3", Last3="Last3", Last3vsFirst3="Last3-First3")
res <- limma_contrasts(M, grp=grp, contrast.v = contr.v)
```

We add a column of gene symbols to the top genes.
```{r sym}
res.df <- data.frame(signif(res, 3), Gene.Symbol=NA)
res.df$Gene.Symbol[1:10] <- LETTERS[1:10]
```

This returns a data frame that looks like:
```{r, echo=FALSE}
knitr::kable(res.df[1:10,])
```

## Significance histogram
We can plot a histogram of the p-values and FDRs for all comparisons with:
```{r sh, eval=FALSE}
signif_hist(res, name=NA) #not run
```

`signif_hist` plots four histograms per page, and this line will plot two histograms per comparison x 3 comparisons = 6 histograms. Thus, this takes more than one page, so it is not shown here.

## Venn diagram
We may want to see which genes are up or down in each comparison, and how the significant genes (say p-value < 0.05) behave in different comparisons using a Venn diagram.

```{r venn}
ezvenn(tab=res, p.cutoff = 0.05)
```

## Dotplot and boxplot
We can plot the top gene by group in a dotplot or boxplot using **ggplot2**. 
```{r pbg}
plot_by_grp(M[1,], pheno$grp, name=NA, type="dot")
plot_by_grp(M[1,], pheno$grp, name=NA, type="box")
```

We can plot multiple genes, such as the first 5 genes where each gene will be on its own page in a PDF, by passing `object=M[1:5,]`.

## Heatmap
We plot a heatmap of the expression of the first five genes across all samples using the **pheatmap** package. We assume their gene symbols are "A", "B", "C", "D", "E". We make a phenotype dataframe with only one column for the group, and with \code{rownames(pheno)==colnames(M)}, which **pheatmap** checks for.

```{r heat}
sym <- c("A", "B", "C", "D", "E")
pheno.df <- pheno[,"grp",drop=F]
topgenes <- rownames(res.df)[1:5]
ezheat(M[topgenes,], labrows = res.df[topgenes, "Gene.Symbol"], pheno.df=pheno.df, name=NA)
```

We see a color bar for the phenotype groups and the strongest change in metabolite "A".

## Volcano plots
We can create volcano plots, which have logFC on the x-axis and log(p-value) on the y-axis, for a comparison using **ggplot2**.
```{r vol}
ezvolcano(tab=res.df, comparison = "First3", name=NA, lab.col="Gene.Symbol", ntop.sig = 1)
```

We can automatically create a PDF with volcano plots for all comparisons with:
```{r mvol, eval=FALSE}
multi_volcano(tab=res.df, lab.col="Gene.Symbol", ntop.sig = 1)
```

## System info
Here is the information on the system on which this document was compiled:
```{r si}
sessionInfo()
```
